A statistical framework for analysing whole-genome bisulfite-sequencing data

### 2015-05-04 Rmarkdown/appendix.Rmd
Tick off a bunch of TODOs

- `methtuple` (`v1.4.0`)
- `GenomicTuples` (`v1.2.1`)
- `MethylationTuples` (`v0.3.0.9007`)
- `methsim` (`v0.5.0.9013`)

@@

```{r R_packages, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
library(xtable)
x <- read.table("../data/r_packages.txt",  header = TRUE,
                stringsAsFactors = FALSE, sep = "\t")
# Drop packages with "*" column set, which indicates the package is a
# "dependency of a dependency".
x <- x[x[, 2] != "*", ]
x <- x[, -2]
xt <- xtable(x, caption = c("R packages used in thesis as reported by \texttt{devtools::session\\_info()}", "R packages used in thesis"),
             label = "tab:R_packages")
align(xt) <- rep("c", ncol(x) + 1)
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable",
      floating = FALSE, include.rownames = FALSE, caption.placement = "top")
```

>>>>

- `methtuple` (`v1.4.0`)
- `GenomicTuples` (`v1.2.1`)
- `MethylationTuples` (`v0.3.0.9007`, commit `4e127d2`)
- `methsim` (`v0.5.0.9013`, commit `9162c8b`)

@@

```{r R_packages, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
library(xtable)
x <- read.table("../data/r_packages.txt",  header = TRUE,
                stringsAsFactors = FALSE, sep = "\t")
# Drop packages with "*" column set, which indicates the package is a
# "dependency of a dependency".
x <- x[x[, 2] != "*", ]
x <- x[, -c(2, 4)]
# Drop MethylationTuples and methsim which are reported elsewhere
x <- x[!x$package %in% c("methsim", "MethylationTuples", "GenomicTUples"), ]
xt <- xtable(x, caption = c("R packages used in thesis as reported by \\texttt{devtools::session\\_info()}", "R packages used in thesis"),
             label = "tab:R_packages")
align(xt) <- rep("l", 4)
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable",
      floating = FALSE, include.rownames = FALSE, caption.placement = "top")
```

### 2015-05-04 Rmarkdown/methsim.Rmd
Tick off a bunch of TODOs

While appropriate for comparing alignment strategies, `Sherman` produces data that is not suitable for use in comparing downstream analysis methods. All CG (resp. CH) loci have an average $\beta$-value of `-CG` (resp. `-CG`) regardless of the genomic context, which we know to be incompatible with real data. Furthermore, the methylation state of each methylation locus is independent, which is clearly inconsistent with the strong co-methylation observed in real data.

@@

Once we have a true methylome, we can simulate data from it. `methsim` currently supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method. Other assays, such as RRBS or microarrays could in principle be supported. A `WGBSParam` object will contain a `SimulatedMethylome` object, along with parameters such as the read-length, sequencing coverage and error rate of the data to be simulated. When applied to a `WGBSParam` object, the `simulate()` method returns a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2]. A `SimulatedBS` object contains all simulated reads and is generally a large object (on the order of $10$ GB). By contrast, the `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat], is much smaller (on the order of $100$ MB) but does not contain the full information of the simulation since read-level data are lost. `methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.

@@

\subsubsection{Simulating differential methylation}\label{sec:sim_diff_meth}

@@

\subsubsection{Performance}

While `methsim` is an R package, much of it is written in C++, using the Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures. Furthermore, several steps of the simulation can be run in parallel for each chromosome. Parallelism is implemented via the `BiocParallel` Bioconductor package \citep{BiocParallel}.

@@

Most notably, and frustratingly, I do not yet have an adequate model for the correlations of the true methylation levels, $\mathbf{B}$. The model of \citet{Lacey:2013iy} is promising, although may be computationally infeasible for whole-genome data given the timings reported in the original publication. The model used by `WGBSsuite` \citet{Rackham:2015bv} is similar to m2, so I suspect it will produce similar results, but this requires further investigation. I continue to explore how alternative models of $\mathbf{B}$ might be integrated into `methsim` to improve this aspect of the simulation.

>>>>

While appropriate for comparing alignment strategies, `Sherman` produces data that is not suitable for use in comparing downstream analysis methods. All CG (resp. CH) loci have an average $\beta$-value of `-CG` (resp. `-CH`) regardless of the genomic context, which we know to be incompatible with real data. Furthermore, the methylation state of each methylation locus is independent, which is clearly inconsistent with the strong co-methylation observed in real data.

@@

Once we have a true methylome, we can simulate data from it. `methsim` currently supports the simulation of whole-genome bisulfite-sequencing reads via the `WGBSParam` class and associated `simulate()` method. Other assays, such as RRBS or microarrays could in principle be supported. A `WGBSParam` object will contain a `SimulatedMethylome` object, along with parameters such as the read-length, sequencing coverage and error rate of the data to be simulated. When applied to a `WGBSParam` object, the `simulate()` method returns a `SimulatedBS` object or a `MethylationTuples::MethPat` object[^namespace2]. A `SimulatedBS` object contains all simulated reads and is generally a large object (on the order of $10$ GB). By contrast, the `MethylationTuples::MethPat` object, which summarises the simulated reads for m-tuples of a particular size[^methpat], is much smaller (on the order of $500$ MB) but does not contain the full information of the simulation since read-level data are lost. `methsim` also includes a helper function, `asMethPat()`, to coerce a `SimulatedBS` object to a `MethylationTuples::MethPat` object.

@@

\subsection{Simulating differential methylation}\label{sec:sim_diff_meth}

@@

\subsection{Performance}

While `methsim` is an R package, much of it is written in \CC, using the Rcpp package \citep{Eddelbuettel:2011to, Eddelbuettel:2013if}, which greatly speeds up the running time of key procedures. Furthermore, several steps of the simulation can be run in parallel for each chromosome. Parallelism is implemented via the `BiocParallel` Bioconductor package \citep{BiocParallel}.

@@

Most notably, and frustratingly, I do not yet have an adequate model for the correlations of the true methylation levels, $\mathbf{B}$. The model of \citet{Lacey:2013iy} is promising, although may be computationally infeasible for whole-genome data given the timings reported in the original publication ($13$ seconds for $5,000$ CpGs increasing to $80$ seconds for $10,000$ CpGs). The model used by `WGBSsuite` \citet{Rackham:2015bv} is similar to m2, so I suspect it will produce similar results, but this requires further investigation. I continue to explore how alternative models of $\mathbf{B}$ might be integrated into `methsim` to improve this aspect of the simulation.

### 2015-05-04 Rmarkdown/wgbs_downstream_analyses.Rmd
Tick off a bunch of TODOs

`MethPipe` is perhaps the best documented and potentially extensible software for analysing methylation patterns at m-tuples. `MethPipe` is mostly written in C++ and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data. As such, it does not feature tools that are particularly amenable to interactive or exploratory analyses. In fact, I am unaware of any software that allows easy exploratory analyses of methylation patterns at m-tuples, which in part motivated the development of `MethylationTuples`.

@@

`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples. It is released under an Artistic-2.0 license, consistent with core Bioconductor packages. I chose to write this software in R because it is a very popular language for data analysis, particularly in bioinformatics, and facilitates both batch and interactive usage. R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code. To improve the performance of key functionality, parts of `MethylationTuples` are written in `C++`, making use of the `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.

>>>>

`MethPipe` is perhaps the best documented and potentially extensible software for analysing methylation patterns at m-tuples. `MethPipe` is mostly written in \CC and is designed as a suite of tools for a complete 'pipeline' analysis of bisulfite-sequencing data. As such, it does not feature tools that are particularly amenable to interactive or exploratory analyses. In fact, I am unaware of any software that allows easy exploratory analyses of methylation patterns at m-tuples, which in part motivated the development of `MethylationTuples`.

@@

`MethylationTuples` is an R package for managing, analysing and visualising methylation patterns at m-tuples. It is released under an Artistic-2.0 license, consistent with core Bioconductor packages. I chose to write this software in R because it is a very popular language for data analysis, particularly in bioinformatics, and facilitates both batch and interactive usage. R is also my computational mother tongue and an R package is a convenient unit for sharing reusable code. To improve the performance of key functionality, parts of `MethylationTuples` are written in \CC, making use of the `Rcpp` package \citep{Eddelbuettel:2011to,Eddelbuettel:2013if}.

### 2015-05-04 Rmarkdown/wgbs_statistical_framework.Rmd
Tick off a bunch of TODOs

```{r EPISCOPE_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
library(xtable)
x <- read.table("../data/strand_correlation/EPISCOPE_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{EPISCOPE} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{EPISCOPE}: Correlations of $\\beta$-values across strands"), label = "tab:EPISCOPE_strand_correlations")
align(xt) <- rep("c", ncol(x) + 1)
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = "top")
```

```{r Lister_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Lister_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Lister} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Lister}: Correlations of $\\beta$-values across strands"), label = "tab:Lister_strand_correlations")
align(xt) <- rep("c", ncol(x) + 1)
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
```

```{r Seisenberger_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Seisenberger_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Seisenberger} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Seisenberger}: Correlations of $\\beta$-values across strands"), label = "tab:Seisenberger_strand_correlations")
align(xt) <- rep("c", ncol(x) + 1)
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
```

```{r Ziller_merged_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Ziller_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Ziller} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Ziller}: Correlations of $\\beta$-values across strands"), label = "tab:Ziller_merged_strand_correlations")
align(xt) <- rep("c", ncol(x) + 1)
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
```

>>>>

```{r EPISCOPE_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
library(xtable)
x <- read.table("../data/strand_correlation/EPISCOPE_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
x <- x[, 1:5]
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{EPISCOPE} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{EPISCOPE}: Correlations of $\\beta$-values across strands"), label = "tab:EPISCOPE_strand_correlations")
align(xt) <- c("r", rep("c", ncol(x)))
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = "top")
```

```{r Lister_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Lister_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
x <- x[, 1:5]
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Lister} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Lister}: Correlations of $\\beta$-values across strands"), label = "tab:Lister_strand_correlations")
align(xt) <- c("r", rep("c", ncol(x)))
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
```

```{r Seisenberger_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Seisenberger_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
x <- x[, 1:5]
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Seisenberger} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Seisenberger}: Correlations of $\\beta$-values across strands"), label = "tab:Seisenberger_strand_correlations")
align(xt) <- c("r", rep("c", ncol(x)))
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
```

```{r Ziller_merged_strand_correlations, echo = FALSE, eval = TRUE, message = FALSE, results = "asis"}
x <- read.table("../data/strand_correlation/Ziller_merged_strand_correlations.txt",
stringsAsFactors = FALSE)
colnames(x) <- 1:10
x <- x[, 1:5]
xt <- xtable(x, caption = c("Correlations of $\\beta$-values across strands for the \\emph{Ziller} dataset. Each row is a sample and each column is a minimum coverage cutoff.", "\\emph{Ziller}: Correlations of $\\beta$-values across strands"), label = "tab:Ziller_merged_strand_correlations")
align(xt) <- c("r", rep("c", ncol(x)))
digits(xt) <- 2
print(xt, booktabs = TRUE, comment = FALSE, tabular.environment = "longtable", floating = FALSE, caption.placement = 'top')
```
